Halving the cost of quantum addition

### 2018-01-19 paper.tex
Re-insert figures into main text flow
Quantum style

\documentclass[twocolumn]{revtex4-1}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{float}
\usepackage[pdfpagelabels,pdftex,bookmarks,breaklinks]{hyperref}
\usepackage{tikz}
\usepackage[all]{hypcap}
\hypersetup{colorlinks,citecolor=blue,urlcolor=blue,linkcolor=blue}

@@

\begin{abstract}
We improve the number of T gates needed to perform an $n$-bit adder from $8n + O(1)$ \citep{Amy2013, Cuccaro2004, AustinDiscussionsAndEmails2017} to $4n + O(1)$.
We do so via a ``temporary logical-AND" construction which uses four T gates to store the logical-AND of two qubits into an ancilla and zero T gates to later erase the ancilla.
This construction is equivalent to one by Jones \cite{Jones2013}, except that our framing makes it clear that the technique is far more widely applicable than previously realized.
Temporary logical-ANDs can be applied to integer arithmetic, modular arithmetic, rotation synthesis, the quantum Fourier transform, Shor's algorithm, Grover oracles, and many other circuits.
Because T gates dominate the cost of quantum computation based on the surface code, and temporary logical-ANDs are widely applicable, this represents a significant reduction in projected costs of quantum computation.
In addition to our $n$-bit adder, we present an $n$-bit controlled adder circuit with T-count of $8n + O(1)$, a temporary adder that can be computed for the same cost as the normal adder but whose result can be kept until it is later uncomputed without using T gates, and discuss some other constructions whose T-count is improved by the temporary logical-AND.
\end{abstract}

@@

The surface code \citep{Brav98,Denn02,Raus07,Raus07d,Fowler2012} is a quantum error correcting code that works on a 2D nearest-neighbour array of qubits and achieves a threshold error rate of approximately 1\%.
This makes the surface code a likely component in the architecture of future error corrected quantum computers, because 2D arrays of qubits with nearest-neighbor connections are possible with many qubit technologies \citep{Schl11,Bare13,Gamb17,Leik17,Laht17} and other well understood error correcting codes either have lower thresholds or require stronger connectivity.

@@

The textbook construction of a Toffoli gate uses seven T gates \citep{Nielsen2009}.
When Toffoli operations are paired, i.e. when an initial Toffoli operation is later uncomputed by a second Toffoli operation, each Toffoli in the pair can omit three of the T gates from the textbook construction.
This introduces phase errors but, assuming intermediate operations aren't sensitive to the phase errors, the second Toffoli gate can uncompute the phase errors while uncomputing the state permutation \citep{Barenco1995, Nielsen2009}.
It is also possible to reduce the T-count of an unpaired Toffoli gate to 4 by using an ancilla qubit and a classically conditioned fixup operation \citep{Jones2013}.

% [figure 1 "adder" was here]

The Cuccaro adder uses $2n + O(1)$ Toffoli gates \citep{Cuccaro2004}.
Existing constructions implement the Cuccaro adder's Toffolis using $8n + O(1)$ T gates \citep{Barenco1995, Cuccaro2004, Amy2013}.
This T-count can be achieved either by applying the ancilla-and-fixup construction from \citep{Jones2013} to each individual Toffoli gate, or by noting that all but one of the adder's Toffoli gates appear in compute/uncompute pairs and then applying the matched-phase-error construction from \citep{Barenco1995, Nielsen2009} to each pair.

The leading factor of 8 in the T-count of an $n$-bit adder has stood for over a decade \citep{Barenco1995, Cuccaro2004, AustinDiscussionsAndEmails2017}.
We improve the leading factor with a construction based on the temporary logical-AND, combining ideas from both the matched-phase-error construction and the ancilla-and-fix-up construction.
We thereby halve the number of T gates needed to perform Toffoli gates that appear in compute/uncompute pairs and thus halve the T-count of Cuccaro-style adders from $8n + O(1)$ to $4n + O(1)$.

@@

% [figure 2 "adder-building-block" was here]

@@

Some additions, such as the ones performed by the multiplications within the modular exponentiation in Shor's algorithm, are conditioned on a control qubit.
Our controlled adder reduces the cost of these additions from $21n + O(1)$ \citep{Coreas2017} to $8n + O(1)$.

@@

% [figure 3 "temporary-logical-and" was here]

Decreasing the T-count of addition reduces the T-count of any construction based on addition.
For example, in \citep{Fowler2012} it is estimated that factoring a 2048-bit number on a surface-code-based quantum computer would take 27 hours and $2 \cdot 10^{12}$ distilled $|T\rangle$ states.
The time estimate is based on Toffolis having a T-depth of 3, and the $|T\rangle$ count estimate is based on Toffolis having a T-count of 7.
Because Shor's algorithm is dominated by the cost of controlled additions, our techniques (and previous work) reduce the average T-count and T-depth of its Toffolis to $\sim 2.7$ and $1$ respectively.
This reduces the time estimate to 9 hours and the count estimate to $8 \cdot 10^{11}$ distilled $|T\rangle$ states.

@@

% [figure 4 "other-adder-building-blocks" was here]

@@

% [figure 5 "paired-toffoli-replacement-rule" was here]

@@

Yet another example of a kind of circuit that benefits from the temporary logical-AND is {\em low-depth} adder circuits.
For example, the Toffoli gates produced by the $P$ and $P^{-1}$ rounds in Draper et al's logarithmic-depth adder \cite{Draper2004} form compute/uncompute pairs that can be replaced by temporary logical-ANDs.
Although using temporary logical-ANDs improves the T-count of logarithmic-depth adders, the resulting T-count is still worse than the T-count of linear-depth adders.
This means that in practice, given our expectation that error corrected quantum computation will bottleneck waiting on $|T\rangle$ states to perform T gates, the linear-depth adders will often execute in less time than the logarithmic-depth adders.
In fact, by using temporary logical-ANDs to implement the classical Brent-Kung adder \cite{brent1982} in a reversible fashion, two $N$-bit numbers can be added in $O(\lg N)$ depth with a T-count of only $12N$.

Finally, we note that Toffolis also appear in compute/uncompute pairs in quantum circuits rooted in physics instead of mathematics.
For example, we used temporary logical-ANDs to cut the projected T-count of a step in a chemistry algorithm under development by nearly 30\% \citep{RyanEmails2017}.

@@

For over a decade, the T-count of addition has been $8n + O(1)$ \citep{Amy2013, Barenco1995, Cuccaro2004}.
In this paper we showed how to halve the leading factor of this cost by replacing Toffolis in compute/uncompute pairs with temporary logical-ANDs.
The temporary logical-AND is a basic circuit construction widget that can be used in many circuits, and a good example of measurement breaking the symmetry between computation and uncomputation (allowing one to be more efficient than the other).
We demonstrated how to optimize the T-count of a few tasks using our adder and the temporary logical-AND, but there are many other low hanging applications (e.g. converting between binary and unary, temporary sorting, applying an operation to a qubit indexed by a register in superposition, parameterized bit-rotation of a qubit register, computing the greatest common divisor of two values, etc).

@@

% [figure 6 "synthesize-t-gates-from-temporary-logical-and" was here]

@@

\section*{Methods}

Not applicable. The circuits can be verified with any quantum circuit simulator.

@@

\section*{Competing Interests}

None.

\section*{Contributions}

Craig Gidney: sole author.

\section*{Funding}

None. Work was done while employed at Google.

@@

\section*{Figure legends}

None.

\section*{Tables}

None.

\section*{Figures}

\begin{figure}[H]
  \includegraphics[width=\linewidth]{adder.pdf}
  \caption{
	A 5-bit adder with T-count of 16.
	Uses Clifford operations, four logical-AND computations each with a T-count of 4, and four logical-AND erasures requiring no T gates.
	Generalizes to an $n$-bit adder with a T-count of $4n - 4$.
	See \autoref{fig:adder-building-block} for the adder building-block and \autoref{fig:temporary-logical-AND} for the logical-AND computation and uncomputation circuits.
  }
  \label{fig:adder}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\linewidth]{adder-building-block.pdf}
  \caption{
	Our adder circuit building-block, with a T-count of 4.
	A variant of the Cuccaro adder building-block \citep{Cuccaro2004}.
  }
  \label{fig:adder-building-block}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\linewidth]{temporary-logical-and.pdf}
  \caption{
	How to compute and uncompute the logical-AND of two qubits.
	The computation circuit (top) has a T-count of 4 and a T-depth of 1.
	Note that the $|T\rangle$ state input contributes to the T-count, because $|T\rangle$ states are the resource used to perform T gates.
	The uncomputation circuit (bottom) uses a measure-and-fixup approach \cite{Jones2013} that requires only Clifford gates, and so has a T-count of zero.
    \\
    An alternative uncomputation construction is to simply do the reverse of the computation circuit.
    This alternative approach has a net T-count of 2 (because an $|T\rangle$ state is recovered).
    The resulting temporary logical-AND with a T-count of 6 would still be an improvement on existing work, but would be inferior to the measure-and-fixup approach shown above.
  }
  \label{fig:temporary-logical-AND}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\linewidth]{other-adder-building-blocks.pdf}
  \caption{
	Variations on our adder construction.
	(a) is a controlled-adder building-block with a T-count of 8.
	The addition will only occur when the control is on.
	(b) is a temporary-adder building-block with a T-count of 4.
	Has the same T-count as a normal addition, but makes the result available before any uncomputation steps are needed.
  }
  \label{fig:other-adder-building-blocks}
\end{figure}

\begin{figure}[H]
  \includegraphics[width=\linewidth]{paired-toffoli-replacement-rule.pdf}
  \caption{
	A sufficient condition for replacing a pair of Toffoli gates with a temporary logical-AND, saving 4 T gates.\\
	1) The later Toffoli gate must be uncomputing the earlier Toffoli gate.\\
	2) Intermediate operations must not be sensitive to the presence of the entangled ancilla.
  }
  \label{fig:paired-toffoli-replacement-rule}
\end{figure}

\vspace{5cm}

\begin{figure}[H]
  \includegraphics[width=\linewidth]{synthesize-t-gates-from-temporary-logical-and.pdf}
  \caption{
	Three T gates can be synthesized by performing a temporary logical-AND, one T gate, and various Clifford operations.
	Clifford operations can't produce T gates, so the extra T gates must be coming from the temporary logical-AND.
	This proves that performing the temporary logical-AND requires at least two T gates.
  }
  \label{fig:synthesize-t-gates-from-temporary-logical-and}
\end{figure}

>>>>

\documentclass[twocolumn]{quantumarticle-customized}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{float}
\usepackage[pdfpagelabels,pdftex,bookmarks,breaklinks]{hyperref}
\usepackage{tikz}
\usepackage[all]{hypcap}
\hypersetup{colorlinks,citecolor=blue,urlcolor=blue,linkcolor=blue}

@@

\begin{abstract}
We improve the number of T gates needed to perform an $n$-bit adder from $8n + O(1)$ \cite{Amy2013, Cuccaro2004, AustinDiscussionsAndEmails2017} to $4n + O(1)$.
We do so via a ``temporary logical-AND" construction which uses four T gates to store the logical-AND of two qubits into an ancilla and zero T gates to later erase the ancilla.
This construction is equivalent to one by Jones \cite{Jones2013}, except that our framing makes it clear that the technique is far more widely applicable than previously realized.
Temporary logical-ANDs can be applied to integer arithmetic, modular arithmetic, rotation synthesis, the quantum Fourier transform, Shor's algorithm, Grover oracles, and many other circuits.
Because T gates dominate the cost of quantum computation based on the surface code, and temporary logical-ANDs are widely applicable, this represents a significant reduction in projected costs of quantum computation.
In addition to our $n$-bit adder, we present an $n$-bit controlled adder circuit with T-count of $8n + O(1)$, a temporary adder that can be computed for the same cost as the normal adder but whose result can be kept until it is later uncomputed without using T gates, and discuss some other constructions whose T-count is improved by the temporary logical-AND.
\end{abstract}

@@

The surface code \cite{Brav98,Denn02,Raus07,Raus07d,Fowler2012} is a quantum error correcting code that works on a 2D nearest-neighbour array of qubits and achieves a threshold error rate of approximately 1\%.
This makes the surface code a likely component in the architecture of future error corrected quantum computers, because 2D arrays of qubits with nearest-neighbor connections are possible with many qubit technologies \cite{Schl11,Bare13,Gamb17,Leik17,Laht17} and other well understood error correcting codes either have lower thresholds or require stronger connectivity.

\begin{figure}
  \includegraphics[width=\linewidth]{adder.pdf}
  \caption{
	A 5-bit adder with T-count of 16.
	Uses Clifford operations, four logical-AND computations each with a T-count of 4, and four logical-AND erasures requiring no T gates.
	Generalizes to an $n$-bit adder with a T-count of $4n - 4$.
	See \autoref{fig:adder-building-block} for the adder building-block and \autoref{fig:temporary-logical-AND} for the logical-AND computation and uncomputation circuits.
  }
  \label{fig:adder}
\end{figure}

@@

The textbook construction of a Toffoli gate uses seven T gates \cite{Nielsen2009}.
When Toffoli operations are paired, i.e. when an initial Toffoli operation is later uncomputed by a second Toffoli operation, each Toffoli in the pair can omit three of the T gates from the textbook construction.
This introduces phase errors but, assuming intermediate operations aren't sensitive to the phase errors, the second Toffoli gate can uncompute the phase errors while uncomputing the state permutation \cite{Barenco1995, Nielsen2009}.
It is also possible to reduce the T-count of an unpaired Toffoli gate to 4 by using an ancilla qubit and a classically conditioned fixup operation \cite{Jones2013}.

The Cuccaro adder uses $2n + O(1)$ Toffoli gates \cite{Cuccaro2004}.
Existing constructions implement the Cuccaro adder's Toffolis using $8n + O(1)$ T gates \cite{Barenco1995, Cuccaro2004, Amy2013}.
This T-count can be achieved either by applying the ancilla-and-fixup construction from \cite{Jones2013} to each individual Toffoli gate, or by noting that all but one of the adder's Toffoli gates appear in compute/uncompute pairs and then applying the matched-phase-error construction from \cite{Barenco1995, Nielsen2009} to each pair.

\begin{figure}
  \includegraphics[width=\linewidth]{adder-building-block.pdf}
  \caption{
	Our adder circuit building-block, with a T-count of 4.
	A variant of the Cuccaro adder building-block \cite{Cuccaro2004}.
  }
  \label{fig:adder-building-block}
\end{figure}

The leading factor of 8 in the T-count of an $n$-bit adder has stood for over a decade \cite{Barenco1995, Cuccaro2004, AustinDiscussionsAndEmails2017}.
We improve the leading factor with a construction based on the temporary logical-AND, combining ideas from both the matched-phase-error construction and the ancilla-and-fix-up construction.
We thereby halve the number of T gates needed to perform Toffoli gates that appear in compute/uncompute pairs and thus halve the T-count of Cuccaro-style adders from $8n + O(1)$ to $4n + O(1)$.

@@

@@

Some additions, such as the ones performed by the multiplications within the modular exponentiation in Shor's algorithm, are conditioned on a control qubit.
Our controlled adder reduces the cost of these additions from $21n + O(1)$ \cite{Coreas2017} to $8n + O(1)$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{temporary-logical-and.pdf}
  \caption{
	How to compute and uncompute the logical-AND of two qubits.
	The computation circuit (top) has a T-count of 4 and a T-depth of 1.
	Note that the $|T\rangle$ state input contributes to the T-count, because $|T\rangle$ states are the resource used to perform T gates.
	The uncomputation circuit (bottom) uses a measure-and-fixup approach \cite{Jones2013} that requires only Clifford gates, and so has a T-count of zero.
    \\
    An alternative uncomputation construction is to simply do the reverse of the computation circuit.
    This alternative approach has a net T-count of 2 (because an $|T\rangle$ state is recovered).
    The resulting temporary logical-AND with a T-count of 6 would still be an improvement on existing work, but would be inferior to the measure-and-fixup approach shown above.
  }
  \label{fig:temporary-logical-AND}
\end{figure}

Decreasing the T-count of addition reduces the T-count of any construction based on addition.
For example, in \cite{Fowler2012} it is estimated that factoring a 2048-bit number on a surface-code-based quantum computer would take 27 hours and $2 \cdot 10^{12}$ distilled $|T\rangle$ states.
The time estimate is based on Toffolis having a T-depth of 3, and the $|T\rangle$ count estimate is based on Toffolis having a T-count of 7.
Because Shor's algorithm is dominated by the cost of controlled additions, our techniques (and previous work) reduce the average T-count and T-depth of its Toffolis to $\sim 2.7$ and $1$ respectively.
This reduces the time estimate to 9 hours and the count estimate to $8 \cdot 10^{11}$ distilled $|T\rangle$ states.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{other-adder-building-blocks.pdf}
  \caption{
	Variations on our adder construction.
	(a) is a controlled-adder building-block with a T-count of 8.
	The addition will only occur when the control is on.
	(b) is a temporary-adder building-block with a T-count of 4.
	Has the same T-count as a normal addition, but makes the result available before any uncomputation steps are needed.
  }
  \label{fig:other-adder-building-blocks}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{paired-toffoli-replacement-rule.pdf}
  \caption{
	A sufficient condition for replacing a pair of Toffoli gates with a temporary logical-AND, saving 4 T gates.\\
	1) The later Toffoli gate must be uncomputing the earlier Toffoli gate.\\
	2) Intermediate operations must not be sensitive to the presence of the entangled ancilla.
  }
  \label{fig:paired-toffoli-replacement-rule}
\end{figure}

@@

Yet another example of a kind of circuit that benefits from the temporary logical-AND is {\em low-depth} adder circuits.
For example, the Toffoli gates produced by the $P$ and $P^{-1}$ rounds in Draper et al's logarithmic-depth adder \cite{Draper2004} form compute/uncompute pairs that can be replaced by temporary logical-ANDs.
Alternatively, by using temporary logical-ANDs to implement the classical Brent-Kung adder \cite{brent1982} in a reversible fashion, two $N$-bit numbers can be added in $O(\lg N)$ depth with a T-count of only $12N$.

Finally, we note that Toffolis also appear in compute/uncompute pairs in quantum circuits rooted in physics instead of mathematics.
For example, we used temporary logical-ANDs to cut the projected T-count of a step in a chemistry algorithm under development by nearly 30\% \cite{RyanEmails2017}.

\begin{figure}
  \includegraphics[width=\linewidth]{synthesize-t-gates-from-temporary-logical-and.pdf}
  \caption{
	Three T gates can be synthesized by performing a temporary logical-AND, one T gate, and various Clifford operations.
	Clifford operations can't produce T gates, so the extra T gates must be coming from the temporary logical-AND.
	This proves that performing the temporary logical-AND requires at least two T gates.
  }
  \label{fig:synthesize-t-gates-from-temporary-logical-and}
\end{figure}

@@

For over a decade, the T-count of addition has been $8n + O(1)$ \cite{Amy2013, Barenco1995, Cuccaro2004}.
In this paper we showed how to halve the leading factor of this cost by replacing Toffolis in compute/uncompute pairs with temporary logical-ANDs.
The temporary logical-AND is a basic circuit construction widget that can be used in many circuits, and a good example of measurement breaking the symmetry between computation and uncomputation (allowing one to be more efficient than the other).
We demonstrated how to optimize the T-count of a few tasks using our adder and the temporary logical-AND, but there are many other low hanging applications (e.g. converting between binary and unary, temporary sorting, applying an operation to a qubit indexed by a register in superposition, parameterized bit-rotation of a qubit register, computing the greatest common divisor of two values, etc).

@@

@@

@@

@@
